---
title: "Getting started with XeniumClean"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with XeniumClean}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This vignette walks through a complete XeniumClean workflow. The example uses
Xenium nomenclature throughout but the package works equally well with
MERSCOPE, CosMx, or any other imaging-based spatial dataset stored in a Seurat
object.

## Background

Imaging-based spatial transcriptomics platforms detect single transcripts at
high spatial resolution, but cell segmentation is imperfect. Transcripts can
"leak" into neighbouring cells, especially in densely packed tissue. This
contamination produces cells that appear to express genes characteristic of
their neighbours rather than their own type.

XeniumClean uses a single-cell RNA-seq reference to define, per cell type,
which genes are plausibly expressed and which are not. For every spatial cell,
it then asks: do any of my neighbours belong to a cell type that genuinely
expresses gene X, while my own cell type does not? If so, gene X in my count
vector is contamination and is set to zero.

```
+--- Cell A (T cell) ----------+
|                              |
|  Genuine T cell genes: kept  |
|  Macrophage neighbour genes  |
|  that T cells do not express:|
|     erased                   |
+------------------------------+
```

Cleanup is conservative. Genes that A could plausibly express are never
removed, even if A's neighbours also express them.

# Critical considerations — read before running

## 1. Cell types must be pre-defined in both datasets

XeniumClean does not perform clustering or annotation. Both the reference and
your spatial object need a populated cell type column before you start. Label
names must match between reference and spatial — use `CollapseLabels()` to
align them if needed.

## 2. Use BROAD cell types, not fine subsets

Define cell types at the **major lineage level**:

✅ Good cell type categories:

```
T_cells, B_cells, NK_cells, Plasma_cells, Macrophage, Monocytes, DC,
Mast_cells, Neutrophils, Cancer, Fibroblasts, Endothelial, PVL
```

❌ Avoid splitting too finely for cleanup:

```
CD4_T_cells, CD8_T_cells, T_reg_cells, Cycling_T_cells,
Memory_B_cells, Naive_B_cells,
M1_Macrophage, M2_Macrophage, TREM2_Macrophage,
cDC1, cDC2, pDC,
Cancer_Basal, Cancer_LumA, Cancer_LumB, ...
```

Why broad:

- Closely related subsets share most of their expression program. If you split
  T cells into CD4 vs CD8, real T cell genes that happen to be slightly higher
  in one will be in the other's `NotExpressedGenes` and get erased from real
  cells.
- The reference must have enough cells per group to estimate gene fractions
  stably (≥30 per group, ideally ≥100). Fine subsets often fall below this.
- Differences between fine subsets are exactly what you want to discover
  *after* cleanup, in cleaner expression space.

Even consider **grouping NK cells with T cells** (as "lymphoid") if your panel
can't reliably tell them apart. The reverse — splitting T cells into CD4/CD8
for cleanup — is almost always a mistake.

## 3. Subclustering after cleanup reveals real biology

This is a significant secondary benefit of XeniumClean.

In raw Xenium data, fine subclustering of e.g. T cells often picks up
**neighbourhood patterns** rather than true T cell biology. T cells near
macrophages contain bled-through macrophage transcripts. T cells near B cells
contain bled-through B cell transcripts. When you cluster these T cells at
high resolution, the clusters reflect **where the T cells live in tissue**,
not what state they're in.

After XeniumClean:

- Macrophage genes are removed from T cells next to macrophages
- B cell genes are removed from T cells next to B cells
- What remains is the T cell's genuine expression

When you now subcluster the cleaned T cells, the clusters reflect **real T
cell biology**: CD4 vs CD8, naive vs effector vs memory, cytotoxic vs
exhausted, etc.

**Recommended workflow:**

```
1. Cluster spatial data broadly → assign broad cell types
2. Run XeniumClean using broad cell types
3. THEN subcluster the cleaned assay to discover true subsets
```

## 4. ⚠️ The reference choice is the single most important decision

XeniumClean is only as good as its reference. A mismatched reference can:

- Mark genes as "not expressed" that ARE expressed in your tissue → those
  genes get erased = **real biology destroyed**
- Mark genes as "expressed" when they shouldn't be → contaminating
  transcripts kept = no cleanup happens

### Reference quality, in order of safety:

| Tier | Reference type | Notes |
|------|----------------|-------|
| ✅✅ Best | scRNA-seq from the same experiment | Matched tissue, samples, sample prep, disease state |
| ✅ Good | Published atlas of the same tissue + disease | e.g. Wu 2021 for breast cancer, Pelka 2021 for CRC |
| ⚠️ Risky | Published atlas of healthy tissue when you have disease | Misses disease-specific expression programs |
| ⚠️ Risky | Published atlas of a related but different tissue | e.g. PBMC reference for tumor tissue T cells |
| ❌ Don't | Manually curated gene lists from literature | Almost impossible to get both expressed AND not-expressed correct |

### Why user-defined references are risky

You can manually build a `gene.sets` object:

```r
gene_sets <- list(
  ExpressedGenes = list(
    T_cells   = c("CD3D","CD3E","CD2","IL7R","..."),
    B_cells   = c("MS4A1","CD79A","CD19","...")
  ),
  NotExpressedGenes = list(
    T_cells   = c("MS4A1","CD79A","CD68","CD163","..."),
    B_cells   = c("CD3D","CD3E","CD68","CD163","...")
  )
)
class(gene_sets) <- "XeniumCleanGeneSets"
```

But to get this right for a 380-gene panel × 10 cell types, you'd need 3800
correct gene-level decisions about whether each gene is expressed (≥10%) or
not-expressed (≤5%) in each cell type. Almost no one can do this accurately
from literature alone, because:

- Marker genes from literature are usually "expressed at high level in", not
  "expressed in ≥10% of cells"
- The "not expressed" list is what protects real biology — getting one gene
  wrong here means erasing real expression
- Cell-type-specific expression varies enormously by tissue, disease state,
  and species

If you must build a reference manually, validate exhaustively: every gene in
your panel, every cell type, against published expression studies.

## 5. Multi-sample objects must be split into a list

XeniumClean operates per-section because spatial neighbours are
within-section. If you've merged samples into one Seurat object, split first.

**Do not use `subset()`** for v5 objects with split layers — it can be slow,
memory-intensive, and produce inconsistent results.

```{r}
# Recommended for most users
xenium_list <- SplitObject(xenium_merged, split.by = "sample")

# Alternative: subset_opt (community function, optimised for v5 split-layer)
# See https://github.com/satijalab/seurat/issues/2724 for the function source
xenium_list <- lapply(unique(xenium_merged$sample), function(s) {
  subset_opt(xenium_merged, subset = sample == s)
})
names(xenium_list) <- unique(xenium_merged$sample)
```

After splitting, each element should:

- Be a single Seurat object for one section
- Have its own FOV (or spatial reduction, or metadata coordinates)
- Carry the cell type column you'll use for cleanup

# Workflow overview

1. Load a single-cell reference (matched to your tissue/disease as closely as
   possible)
2. Load your spatial Seurat object(s); split if merged
3. Build per-cell-type gene sets at BROAD granularity
4. Align label nomenclature between reference and spatial
5. Run cleanup
6. QC
7. Re-cluster on the cleaned assay to discover true subsets

```{r}
library(XeniumClean)
library(Seurat)
```

# Step 1: Load the single-cell reference

```{r}
sc_reference <- readRDS("path/to/single_cell_reference.rds")
sc_reference <- NormalizeData(sc_reference, verbose = FALSE)

# Confirm the label column exists and uses BROAD categories
table(sc_reference$celltype_broad)
```

If the reference has fine subtypes, collapse them first:

```{r}
sc_reference$xclean_label <- CollapseLabels(
  labels  = sc_reference$celltype_minor,
  mapping = c(
    "B cells Memory" = "B_cells",
    "B cells Naive"  = "B_cells",
    "T cells CD4+"   = "T_cells",
    "T cells CD8+"   = "T_cells",
    "Cycling T-cells" = "T_cells",
    "NK cells"        = "T_cells",       # consider grouping with T cells
    "Macrophage"      = "Macrophage",    # include mapping if name is the same
    "Monocyte"        = "Monocytes",
    "DCs"             = "DC",
    "CAFs myCAF-like" = "Fibroblasts",
    "CAFs MSC iCAF-like" = "Fibroblasts",
    "Cancer LumA SC"  = "Cancer",
    "Cancer LumB SC"  = "Cancer",
    "Cancer Basal SC" = "Cancer",
    "Mature Luminal"  = "Cancer"

  )
)
```

# Step 2: Load and prepare your spatial Seurat object(s)

If you have a single section:

```{r}
xenium_obj <- readRDS("path/to/xenium_seurat.rds")
table(xenium_obj$cell_type)
```

If you have multiple sections in a merged object, split it:

```{r}
xenium_list <- SplitObject(xenium_merged, split.by = "sample")
```

# Step 3: Build per-cell-type gene sets

```{r}
# Use whichever object you have - either single or one of the list
shared_genes <- intersect(rownames(sc_reference), rownames(xenium_obj))
length(shared_genes)

gene_sets <- BuildReferenceGeneSets(
  reference                = sc_reference,
  group.by                 = "xclean_label",
  genes.use                = shared_genes,
  expressed.threshold      = 0.10,
  not.expressed.threshold  = 0.05,
  remove.ubiquitous        = TRUE
)

print(gene_sets)
```

**Sanity-check the gene sets** before moving on. Known cell-type-specific
markers should land in the right buckets:

```{r}
"CD3D"  %in% gene_sets$ExpressedGenes$T_cells       # TRUE
"CD3D"  %in% gene_sets$NotExpressedGenes$Cancer     # TRUE
"MS4A1" %in% gene_sets$ExpressedGenes$B_cells       # TRUE
"MS4A1" %in% gene_sets$NotExpressedGenes$T_cells    # TRUE
"EPCAM" %in% gene_sets$ExpressedGenes$Cancer        # TRUE
"CD163" %in% gene_sets$ExpressedGenes$Macrophage    # TRUE
```

If any of these returns FALSE, your reference or label mapping has an issue.
Fix before cleaning.

# Step 4: Align spatial labels to reference granularity

```{r}
xenium_list <- lapply(xenium_list, function(x) {
  x$xclean_label <- CollapseLabels(
    labels  = x$clusters,
    mapping = c(
      # T cells (all merged)
      CD4_T_cells = "T_cells",
      CD8_T_cells = "T_cells",
      T_reg_cells = "T_cells",
      NK_cells    = "T_cells",         # consider grouping
      # DCs
      cDC1_LAMP3  = "DC",
      cDC1_S100B  = "DC",
      cDC2        = "DC",
      pDC         = "DC",
      # Cancer
      Cancer_LumA   = "Cancer",
      Cancer_Basal  = "Cancer",
      Cancer_CXCL14 = "Cancer",       
      Epithelial_Cancer = "Cancer",
      # B cells
      B_cells   = "B_cells"            # include if name is the same
    )
  )
  x
})
```

Labels not in the mapping (e.g. `Neutrophils`, `Mast_cells`, `Adipocytes`)
pass through. Cell types absent from the reference are silently skipped during
cleanup, preserving their counts. This is the safe default.

# Step 5: Run cleanup

```{r}
xenium_list <- XeniumClean(
  object    = xenium_list,
  gene.sets = gene_sets,
  group.by  = "xclean_label",
  radius    = 50,
  parallel  = TRUE,
  workers   = 6
)
```

## Coordinate format options

By default XeniumClean looks for coordinates in image FOVs (Xenium standard),
then a `spatial` reduction, then metadata. To force one or supply your own:

```{r}
# From a dimensional reduction
XeniumClean(xenium_obj, ..., coords.source = "reduction",
            reduction.name = "spatial")

# From metadata columns named cx, cy
XeniumClean(xenium_obj, ..., coords.source = "metadata",
            coord.cols = c("cx", "cy"))

# From a matrix you computed externally
XeniumClean(xenium_obj, ..., coords.source = "matrix", coords = my_xy_matrix)
```

## Skip-pairs for lineage continuum (advanced)

If you kept finer cell types than recommended and are seeing over-cleaning
between closely-related lineages, you can tell XeniumClean to skip those
pairs:

```{r}
xenium_list <- XeniumClean(
  object     = xenium_list,
  gene.sets  = gene_sets,
  group.by   = "xclean_label",
  skip.pairs = list(
    c("T_cells",   "NK_cells"),
    c("B_cells",   "Plasma_cells"),
    c("Monocytes", "Macrophage")
  )
)
```

The cleaner solution is usually to merge these into broader categories
upstream (see Step 4) — skip.pairs is for cases where you really do want
both subtypes labelled separately.

# Step 6: QC

```{r}
CleanupSummary(xenium_list, group.by = "xclean_label")
```

Expect roughly 2 to 8 % of total transcripts removed per sample for a 300 to
500 gene panel. Much higher than 15 % suggests over-cleaning (likely a label
alignment or granularity issue); much lower than 1 % suggests the gene sets
are too permissive.

```{r}
PlotRemovedTranscripts(xenium_list, group.by = "xclean_label")
```

# Step 7: Subcluster on the cleaned assay (the payoff)

This is where XeniumClean really pays off. Cluster the cleaned data and you
should see subsets emerging based on true biology rather than neighbourhood
contamination.

```{r}
# Re-merge if you want a single object, or work on the per-sample list
xenium_obj <- merge(xenium_list[[1]], xenium_list[-1])
DefaultAssay(xenium_obj) <- "XeniumClean"

xenium_obj <- NormalizeData(xenium_obj)
xenium_obj <- ScaleData(xenium_obj, features = rownames(xenium_obj))
xenium_obj <- RunPCA(xenium_obj, features = rownames(xenium_obj),
                      reduction.name = "pca_clean")
xenium_obj <- FindNeighbors(xenium_obj, reduction = "pca_clean", dims = 1:30,
                             graph.name = c("clean_nn", "clean_snn"))
xenium_obj <- FindClusters(xenium_obj, graph.name = "clean_snn",
                            resolution = 0.8,
                            cluster.name = "clean_clusters")
xenium_obj <- RunUMAP(xenium_obj, reduction = "pca_clean", dims = 1:30,
                       reduction.name = "umap_clean")

DimPlot(xenium_obj, reduction = "umap_clean", group.by = "clean_clusters")
```

Now look at marker genes for the new subclusters. Expect to see real subsets
that you couldn't separate before:

```{r}
# T cell subsets in the cleaned data
Idents(xenium_obj) <- "clean_clusters"
DotPlot(xenium_obj,
        features = c("CD4","CD8A","CD8B","FOXP3","IL2RA","TCF7","IL7R",
                     "CCR7","GZMB","GZMK","PRF1","CXCR3","CXCR5"),
        idents = subset(xenium_obj, subset = xclean_label == "T_cells") |>
                   Idents() |> unique()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The original `Xenium` (or whatever your source assay was called) is preserved
alongside `XeniumClean`. You can switch back at any time to compare cleaned
vs uncleaned results, but for downstream biological analysis, the cleaned
assay is what you want.

# Optional: per-cell QC scoring

`ScoreCellTypes()` and `FlagMismatches()` provide diagnostics for identifying
cells that may be mislabeled (their expression matches a different reference
cell type) or heavily contaminated (their expression matches their assigned
type, but their neighbours don't):

```{r}
xenium_obj <- ScoreCellTypes(xenium_obj, gene_sets, assay = "Xenium")
xenium_obj <- FlagMismatches(xenium_obj, group.by = "xclean_label")

table(xenium_obj$xeniumclean_diagnosis, xenium_obj$xclean_label)
```

These flags are diagnostic, not corrective. XeniumClean does not act on them
automatically.

# Summary

- **Cell types must be pre-defined** in both reference and spatial data.
- **Use broad cell types**, not fine subsets, for the cleanup step.
- **Subclustering after cleanup** reveals true biology, not contamination
  patterns.
- **The reference must match your tissue/disease**. The safest reference is
  matched scRNA-seq from your own experiment.
- **Multi-sample objects must be split into a list** before cleanup. Use
  `SplitObject()` rather than `subset()` for v5 objects.
